Extracting Angular Observables without a Likelihood\\and Applications to Rare Decays

### 2015-03-12 draft.tex
S7 plots
Revert "tiny typo fix" and fix uncertainty caption
This reverts commit 4c2db4ca257b711117ee93f473933fc1f61c92a0.

Conflicts:
	draft.tex

Our initial motivation for studying what we wish to call the
\emph{method of moments} is the determination of angular observables
in the rare FCNC-mediated decay $\bar{B}\to \bar{K}^*(\to
\bar{K}\pi)\ell^+\ell^-$. However, the method we describe in the
following is general and applies to all decay or scattering
processes that can be formulated in terms of an orthogonal basis of angular functions.
We find a previous work \cite{Dighe:1998vk} that
advocates this method chiefly for the determination of angular
observables in non-leptonic $B$ decays but also mentions the
applicability to semileptonic decays. Our aim is to improve upon this
previous work by studying the uncertainties that are introduced by
mismodeling of the angular distribution, and by working out a recipe
to determine and unfold detector effects. The latter is crucial for
the application of the method to real data. We show that
the method of moments has several major advantages over the usual approach
based on likelihood fits:
\begin{enumerate}
    \item Likelihood fits have convergence problems for a small number of
        events, and can require reparametrizations and/or approximations for a
        successful fit to the signal PDF. As an example, see the LHCb
        analysis of the angular distribution in $\bar{B}\to
        \bar{K}^*\ell^+\ell^-$ decays \cite{Aaij:2013qta}.

@@

\begin{figure}[b]
        \centering
            \includegraphics[width=0.45\textwidth]{figs/pull-Q2_5_6_S5_200.png}
            \includegraphics[width=0.45\textwidth]{figs/pull-Q2_5_6_S7_200.png}
        \caption{Pull distribution for the angular observables $S_5$ (upper) and $S_7$ (lower), extracted from $2\cdot 10^5$ studies of
        $200$ simulated events each for the decay $B\to K^*\ell^+\ell^-$. The red curve represents a fit to a Gaussian distribution.}
        \label{fig:pulls}
\end{figure}

@@

    \item We also compare the results as obtained by the method of moments with
        results obtained by a conventional likelihood fit. In particular, we
        study the correlation between the method-of-moment estimators and the
        maximum-likelihood-fit estimators. We run $10^3$ toy analyses, with
        $200$ simulated events per analysis. We show the joint distribution of
        the two estimators in \reffig{correlation}. The two estimators are
        highly correlated. The distribution of the difference of the estimators
        exhibits now bias, which is due to the large number of simulated
        events.  Still, we find that statistical uncertainty on the difference
        of the two estimators is sizeable and can easily become half as large
        as the statistical uncertainty of either estimator.
\end{itemize}
\begin{figure}[t]
        \centering
            \includegraphics[width=0.45\textwidth]{figs/Q2_1_2_S5.pdf}
            \includegraphics[width=0.45\textwidth]{figs/Q2_1_2_S9.pdf}
        \caption{Uncertainty of the angular observables $S_5$ (upper) and $S_9$
            (lower), extracted from $2\cdot 10^5$ studies of simulated events
            for the decay $B\to K^*\ell^+\ell^-$. We show the uncertainty as a
            function of the number of simulated events $N$.  The red curve
        represents a fit to the function given in \refeq{unc-on-mean}.
        Errors correspond to $68\%$ spread of measured uncertainties in the toys.}
        \label{fig:errors}
\end{figure}

@@

First, we find no bias in the determinations of the angular observables even
for a small number of events.  However, due to fewer model assumptions, the
uncertainty on the mean values increases by roughly $10\%$--$30\%$ compared
to likelihood fits.\\
Second, in the absence of detector effects, the method of
moments does not rely on model assumptions for the partial-wave composition
of the PDF. This is explicitly shown for the case of higher
partial waves in multibody final states.\\
Third, we develop a systematic method
for the determination of detector effects that lead to dilution and mixing of
the angular observables. We present an algorithm to calculate the necessary
unfolding matrix, which is only computationally feasible when using the method
of moments.\\
Fourth, the fact that the joint distribution of the angular
observables that results from the method of moments is a multivariate Gaussian
distribution allows for an easy and precise transfer of correlation information
to subsequent theoretical analyses.\\
Last but not least, the resulting
distribution arises without the need for additional constraints in order to
achieve convergence, such as e.g. in some maximum-likelihood fits. Therefore,
results obtained using the method of moments can be directly averaged, without
the need to remove the constraints a posteriori.\\

In conclusion, we argue that the method of moments is a competetive alternative to
maximum-likelihood fits as long as angular distributions are involved. We wish to raise
the interesting prospect of extending this method to applications that feature
PDFs composed from non-angular orthogonal bases.

>>>>

Our initial motivation for studying what we wish to call the
\emph{method of moments} is the determination of angular observables
in the rare FCNC-mediated decay $\bar{B}\to \bar{K}^*(\to
\bar{K}\pi)\ell^+\ell^-$. However, the method we describe in the
following is general and applies to all decay or scattering
processes that can be formulated in terms of an orthogonal basis of angular functions.
We find a previous work \cite{Dighe:1998vk} that
advocates this method chiefly for the determination of angular
observables in non-leptonic $B$ decays but also mentions the
applicability to semileptonic decays. Our aim is to improve upon this
previous work by studying the uncertainties that are introduced by
mismodeling of the angular distribution, and by working out a recipe
to determine and unfold detector effects. The latter is crucial for
the application of the method to real data. We show that
the method of moments has several major advantages over the usual approach
based on likelihood fits:
\begin{enumerate}
    \item Likelihood fits have convergence problems for a small number of
        events, and can require reparametrizations and/or approximations for a
        successful fit to the signal PDF. As an example, see the LHCb
        analysis of the angular distribution in $\bar{B}\to
        \bar{K}^*\mu^+\mu^-$ decays \cite{Aaij:2013qta}.

@@

\begin{figure}[b]
        \centering
            \includegraphics[width=0.45\textwidth]{figs/Q2_5_6_S5_200.pdf}
            \includegraphics[width=0.45\textwidth]{figs/Q2_5_6_S7_200.pdf}
        \caption{Pull distribution for the angular observables $S_5$ (upper) and $S_7$ (lower), extracted from $2\cdot 10^5$ studies of
        $200$ simulated events each for the decay $B\to K^*\ell^+\ell^-$. The red curve represents a fit to a Gaussian distribution.}
        \label{fig:pulls}
\end{figure}

@@

    \item We also compare the results as obtained by the method of moments with
        results obtained by a conventional likelihood fit. In particular, we
        study the correlation between the method-of-moment estimators and the
        maximum-likelihood-fit estimators. We run $10^3$ toy analyses, with
        $200$ simulated events per analysis. We show the joint distribution of
        the two estimators in \reffig{correlation}. The two estimators are
        highly correlated. The distribution of the difference of the estimators
        exhibits now bias, which is due to the large number of simulated
        events.  Still, we find that statistical uncertainty on the difference
        of the two estimators is sizeable and can easily become half as large
        as the statistical uncertainty of either estimator.
\end{itemize}
\begin{figure}[t]
        \centering
            \includegraphics[width=0.45\textwidth]{figs/Q2_1_2_S5.pdf}
            % \includegraphics[width=0.45\textwidth]{figs/Q2_1_2_S7.pdf}
            \caption{Uncertainty $\est{\sigma}_5$ of the angular
              observable $S_5$ extracted from $2\cdot 10^5$ studies of
              simulated events for the decay $B\to
              K^*\ell^+\ell^-$. We show the uncertainty as a function
              of the number of simulated events $N$.  The red curve
              represents a fit to the function given in
              \refeq{unc-on-mean}. The error bars correspond to the
              $68\%$ spread of measured uncertainties in the toys. The
              plot for $\est{\sigma}_7$ is visually indistinguishable
              from the one shown here.}
        \label{fig:errors}
\end{figure}

@@

First, we find no bias in the determinations of the angular
observables even for a small number of events.  However, due to fewer
model assumptions, the uncertainty on the mean values increases by
roughly $10\%$--$30\%$ compared
to likelihood fits.\\
Second, in the absence of detector effects, the method of moments does
not rely on model assumptions for the partial-wave composition of the
PDF. This is explicitly shown for the case of higher
partial waves in multibody final states.\\
Third, we develop a systematic method for the determination of
detector effects that lead to dilution and mixing of the angular
observables. We present an algorithm to calculate the necessary
unfolding matrix, which is computationally feasible only when using
the method
of moments. The algorithm also accounts for higher partial waves.\\
Fourth, the joint distribution of the angular observables resulting
from the method of moments is well approximated by a multivariate
Gaussian distribution even for small number of events $N \sim 30$ both
for the ideal uniform acceptance and a realistic example.  This
facilitates the precise transfer of correlation information
to subsequent theoretical analyses.\\
Last but not least, the resulting distribution arises without the need
for additional model constraints. Thus more observables can be
inferred from the same data than in a likelihood fit. In addition,
the results from the method of moments can be more easily averaged or combined; e.g., in global fits.\\

In conclusion, we argue that the method of moments is a competitive alternative to
maximum-likelihood fits if angular distributions are involved. We wish to raise
the interesting prospect of extending this method to applications that feature
PDFs composed from non-angular orthogonal bases.
